Calibrating MEM seeding heuristics

### 2018-07-19 MEM.tex
Resketched the whole thing with newer theory and Monte Carlo.

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{authblk}
\usepackage[english]{babel}
\usepackage{blkarray}
\usepackage[font=small]{caption}
\usepackage{cite}
\usepackage{graphicx}

@@

\newtheorem{algorithm}{Algorithm}
\newtheorem{corollary}{Corollary}
\newtheorem{definition}{Definition}
%\newtheorem{example}{Example}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
%\newtheorem{remark}{Remark}

@@

\newcommand{\DN}{\scriptstyle{\Downarrow}}
\newcommand{\dn}{\scriptstyle{\downarrow}}
\newcommand{\up}[1]{\scriptstyle{(\uparrow)_{#1}}}
\newcommand{\eq}[1]{\scriptstyle{(\sim)_{#1}}}

\newcommand{\upg}{\scriptstyle{\uparrow}}
\newcommand{\eqg}{\scriptstyle{\sim}}

@@

\author[1,2]{Guillaume J. Filion}
\affil[1]{Genome Architecture, Gene Regulation, Stem Cells and Cancer
Programme, Center for Genomic Regulation (CRG), The Barcelona Institute of
Science and Technology, Dr. Aiguader 88, Barcelona 08003, Spain.}
\affil[2]{University Pompeu Fabra, Doctor Aiguader, 08003 Barcelona,
Spain.}

@@

\begin{abstract}
The abstract will come later.
\end{abstract}

@@

\section{Introduction}
Some introduction

\section{MEM seeds}

\subsection{Seeds}
DEFINE MEM SEED AND SHARED MEM SEED.

EXPLAIN THAT THE TAIL IS AN UNTERMINATED SEGMENT.

\subsection{The seeding alphabet}

In order to compute seeding probabilities, we need to consider that reads
are sequences of letters from the \emph{seeding alphabet} rather than from
the nucleotide alphabet. them as a sequence of symbols telling if the
nucleotides match the target and~/ or the duplicates. Here, we will
consider that reads are sequences of the five symbols $\Downarrow$,
$\downarrow$, $\uparrow$, $\sim$ and $\square$.

The $\Downarrow$ symbol indicates that the nucleotide is not only a
mismatch for the target, but also for every duplicate. The $\downarrow$
symbol indicates that the nucleotide is a mismatch for the target, but is
a match for at least one dupulicate. These two symbols are referred to as
the \emph{error symbols}, because they imply the occurrence of a
sequencing error (this is the only way the read may not match the target).

The remaing three symbols have a more complex usage because they describe
events of the MEM seeding process that are best explained by way of a
concrete example. Fig.~\ref{fig:sketch_MEM} illustrates the case of a read
of size 25 from a sequence with three duplicates. The four files of
squares at the bottom are the target and duplicate sequences, where a
black square means that the decoded nucleotide is a mismatch and a grey
square means that it is a match.

\begin{figure}[h]
\centering
\includegraphics[scale=.9]{sketch_MEM.pdf}
\caption{\textbf{Reads and MEM seeds}. 
Some legend.}
\label{fig:sketch_MEM}
\end{figure}

Now consider streaks of matches until a particular position of the read.
At the 4-th nucleotide, the target scores 4 matches in a row versus 2, 0
and 1 for the duplicates. The target just acquired the strictly longest
streak, which is indicated by the $\uparrow$ symbol. This symbol also
appears at position 17, where the target sequence acquires the strictly
longest streak again (6 matches in a row, versus 4, 0 and 4 for the
dupicates).

At position 23, the target and the first duplicate both acquire the
longest streak (2 matches in a row), which is indicated by the $\sim$
symbol. This symbol is used only at positions where the target
\emph{acquires} the longest streak (together with at least one duplicate).
For instance, it is not used at position 1 nor at position 12 because the
target already had the longest streak (0 in both cases).

Finally, the $\square$ symbol is used for all the other cases,
\textit{i.e.}, when there is no sequencing error, and when the target does
not acquire the longest streak of matches (shared or not with a
duplicate).

Each sequence corresponds to a Bernoulli process where `success' stands
for a match between the sequence and the read, and `failure' stands for a
mismatch. For simplicity, we will refer to the Bernoulli processes as
`threads'.

\begin{definition}
At a given position of the read, a duplicate thread is said to be
\emph{strictly masking} if its success streak is strictly longer than the
match streak of the target. A duplicate thread is said to be
\emph{potentially masking} if its match streak is the same as the target.
\end{definition}

Duplicate threads are thus classified into strictly masking, potentially
masking and not masking, which allows us to describe the usage of the five
symbols.

\begin{figure}[h]
\centering
\includegraphics[scale=.7]{decision_tree.pdf}
\caption{\textbf{Decision tree}. 
Some legend.}
\label{fig:decision_tree}
\end{figure}

We assume that the target sequence has $N$ duplicates that are potential
false positives for the seeding process. We further assume that
duplication was instantaenous and that all $N+1$ sequences diverge
independently from each other at a constant rate. In other words, we
ignore the complications due to the genealogy of the duplication events
and we simply assume that at each position, any given duplicate has the
same nucleotide as the target with probability $1-\mu$. If it does not,
we assume that the duplicate has any of the remaining three
nucleotides with equal probability.

To describe those processes, we will
often use the function $\varphi(n)$ that gives the length of the success
streak at position $n$. The processes are indexed such that $\varphi_0$
refers to the target sequence, and functions $\varphi_j$ $(1 \leq j \leq
N)$ refer to the duplicates. For a concrete example, $\varphi_0(5) = 5$
and $\varphi_1(5) = 3$ in Fig.~\ref{fig:sketch_MEM}.

With these definitions, the decision tree for choosing the next symbol to
insert in a read can be summarized as in Fig.~\ref{fig:decision_tree}.
This sketch will be useful as a reference to compute the probabilities of
the blocks and their weighted generating functions.

\begin{definition}
The main thread is dominant at nucleotide $n$ if $\varphi_0(n) >
\varphi_i(n)$ for all $i$ $(1 \leq i \leq N)$. The main thread is
co-dominant if inequality is not strict and there exists $i$ $(1\leq i
\leq N)$ such that $\varphi_0(n) = \varphi_i(n)$.
\end{definition}

%\begin{figure}[h]
%\centering
%\includegraphics[scale=1]{no_3MEM_graph.pdf}
%\caption{\textbf{Transfer graph of a read without 3-MEM}. 
%Some legend.}
%\label{fig:noMEM_graph}
%\end{figure}

@@

\subsection{Transitions from $\uparrow$}
\label{sec:trans_from_up}

On a $\uparrow$ symbol, the main thread becomes dominant. It will remain
so until the first sequencing error, so the next block will be terminated
by an error symbol (\textit{i.e.} $\Downarrow$ or $\downarrow$).

\begin{definition}
$\omega$ and $\tilde{\omega}$ are the probabilities of the $\Downarrow$
and of the $\downarrow$ error symbols, respectively, \textit{i.e.}
\begin{eqnarray}
\omega &=& p \cdot (1-\mu/3)^N \\
\tilde{\omega} &=& p \cdot \big( 1-(1-\mu/3)^N \big).
\end{eqnarray}
\end{definition}

The weighted generating function of the transition from $(\uparrow)_j$ to
$\Downarrow$ is
\begin{equation}
C^\Downarrow_j(z) = \omega z \sum_{i=0}^{\gamma-j-1} (qz)^i.
\end{equation}

The weighted generating function of the transition from
$(\uparrow)_j$ to $\downarrow$ is
\begin{equation}
C^\downarrow_j(z) = \tilde{\omega} z
  \sum_{i=0}^{\gamma-j-1} (qz)^i.
\end{equation}

\subsection{Transitions from $\Downarrow$}
\label{sec:trans_from_Down}

At a $\Downarrow$ symbol, all the threads have match length $0$ and the
read is `reset' to its initial state. The main thread is co-dominant with
all the alternative threads. The next terminator symbol depends on what
happens first. If all the alternative threads fail before the main thread,
the latter becomes dominant ($\uparrow$ symbol). If not all alternative
threads have failed when the main thread does, we need to distinguish two
cases: if all threads fail together with the main thread, the read is
reset ($\Downarrow$ symbol), otherwise the main read becomes dominated
($\downarrow$ symbol).

If the next terminator is the $\uparrow$ symbol, we need to distinguish
the cases based on the value of $\varphi_0$ (see
section~\ref{sec:trans_from_up}), which is just the position of the
$\uparrow$ symbol after the $\Downarrow$ symbol in this case. A transition
from $\Downarrow$ to $(\uparrow)_j$ consists of a succession of $j$
correct nucleotides -- each with probability $q$. But we must also make
sure that no other $\uparrow$ symbol is inserted before position $j$.

\begin{definition}
$\xi_{i,m}$ is the probability that at least one of $m$ alternative
threads does not fail in $i$ error-free nucleotides, \textit{i.e.}
\begin{equation}
  \xi_{i,m} = 1-(1-(1-\mu)^i)^m.
\end{equation}
\end{definition}

The weighted generating function of the transition from $\Downarrow$ to
$(\uparrow)_i$ is
\begin{equation}
\label{eq:u}
u_i(z) = (\xi_{i-1,N}-\xi_{i,N})(qz)^i.
\end{equation}

\begin{remark}
If $N = 0$, the states $(\downarrow)_2, \ldots, (\downarrow)_{\gamma-1}$
are inacessible. Consistently, the mononomial $u_i(z)$ is equal to $qz$ if
$i = 1$ and $0$ otherwise.
\end{remark}

If the next terminator is $\Downarrow$, there is a limit on the
number of $\square$ symbols. Indeed, if there are $\gamma$ or more
nucleotides between the two $\Downarrow$ symbols, they constitute a shared
MEM seed. As mentioned in section~\ref{sec:trans_from_up}, the weighted
generating function of the $\Downarrow$ symbol is $pz(1-\mu/3)^N$. By the
same rationale as above, the weighted generating function of the
transition from $\Downarrow$ to $\Downarrow$ is
\begin{equation}
A^\Downarrow(z) = \omega z
  \sum_{i=0}^{\gamma-1} \xi_{i,N} \cdot (qz)^i.
\end{equation}

If the next terminator is $\downarrow$ we have to dinstinguish two cases.
In the example below, there is a shared MEM seed only if $\gamma \leq
11$. The main thread is unmasked, so the number of nucleotides between the
error symbols matters.
\begin{inset}
\includegraphics[scale=.9]{ddown_to_down_case_1.pdf}
\end{inset}
In this other example, there can be no MEM seed because the first
duplicate masks the main thread. This depends neither on the value of
$\gamma$ nor on the number of nucleotides between the error symbols.
\begin{inset}
\includegraphics[scale=.9]{ddown_to_down_case_2.pdf}
\end{inset}

So if the $\Downarrow$ and the $\downarrow$ symbols are separated by
$\gamma-1$ or fewer nucleotides, there is no MEM seed and we must only
rule out the presence of other terminators in the segment (in this case
the $\uparrow$ symbol). The weighted generating function is
\begin{equation}
\tilde{\omega} z \sum_{i=0}^{\gamma-1} \xi_{i,N} \cdot (qz)^i.
\end{equation}

On the other hand, if the $\Downarrow$ and the $\downarrow$ symbols are
separated by $\gamma$ or more nucleotides, there could be a shared MEM
seed, so we must enforce the condition that at least one potentially
masking thread masks the main thread up to and including the $\downarrow$
symbol (this incidentally rules out the presence of other terminators in
the segment).

\begin{definition}
$\alpha_i$ is the probability that the first thread has at least one
failure in $i$ error-free nucleotides, followed by a sequencing error,
\textit{i.e.}
\begin{equation}
\alpha_i = 1 - (1-\mu)^i\mu/3
\end{equation}
\end{definition}

The probability that at least one duplicate thread masks the main thread
if the first error occurs at position $i$ is thus $1 - \alpha_{i-1}^N$.
The weighted generatin function is
\begin{equation}
pz\sum_{i=\gamma}^\infty (1 - \alpha_i^N) \cdot (qz)^i.
\end{equation}

Finally, the weighted generating function of the transition from
$\Downarrow$ to $\downarrow$ is
\begin{equation}
A^\downarrow(z) =
\tilde{\omega} z \sum_{i=0}^{\gamma-1} \xi_{i,N} \cdot (qz)^i +
pz\sum_{i=\gamma}^\infty (1 - \alpha_i^N) \cdot (qz)^i.
\end{equation}

\subsection{Transitions from $\downarrow$}
\label{sec:trans_from_down}

>>>>

\usepackage{algorithm2e}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{authblk}
\usepackage[english]{babel}
\usepackage{blkarray}
\usepackage[font=small]{caption}
\usepackage{cite}
\usepackage{graphicx}

\setlength{\thickmuskip}{2mu plus 3mu minus 1mu}
\setlength{\medmuskip}{1mu plus 2mu minus 1mu}

\SetKwComment{Comment}{$\triangleright$\ }{}

@@

%\newtheorem{algorithm}{Algorithm}
\newtheorem{corollary}{Corollary}
\newtheorem{definition}{Definition}
%\newtheorem{example}{Example}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
%\newtheorem{remark}{Remark}

@@

\newcommand{\dn}[1]{\scriptstyle{\downarrow_{/#1}}}
\newcommand{\up}[1]{\scriptstyle{\uparrow_{/#1}}}
\newcommand{\nd}{\scriptstyle{|}}

@@

\author[1,2]{Guillaume J. Filion}
\author[1,2]{Eduard Zorita}
\affil[1]{Genome Architecture, Gene Regulation, Stem Cells and Cancer
Programme, Center for Genomic Regulation (CRG), The Barcelona Institute of
Science and Technology, Dr. Aiguader 88, Barcelona 08003, Spain.}
\affil[2]{University Pompeu Fabra, Doctor Aiguader, 08003 Barcelona,
Spain.}

@@

\begin{abstract}
The abstract will come later.
A popular mapping heuristic is called \emph{seeding}, and a recent variant
is the \emph{MEM seeding} approach. Here we tackle the question of how to
evaluate the probability that MEM seeding hits are on-target. This in turn
expresses the confidence one should have in the location of the mapped DNA
fragment and can be used to calibrate the MEM seeding heuristic, or to
filter the hits at the desired level of confidence.
\end{abstract}

@@

\section{Introduction}

\subsection{Mapping}

Consider the following task referred to as the \emph{true mapping
problem}: a random DNA fragment from a known genome is sequenced with an
imperfect instrument; can we identify the location of the fragment in the
genome?

It seems that the answer depends only on the size of the fragment and the
error rate of the instrument, but there is another important factor to
consider. Most genomes contain repeated sequences. Intuitively, the
problem is more difficult if the random DNA fragment is duplicated. If one
of the duplicates is exactly identical to the original, then the DNA
fragment cannot be mapped with certainty. Even if the duplicates are all
different from the original (which happens because of mutations in the
genome), the read may still be closer to one of the duplicates because of
sequencing errors. So we can never be sure that the read is mapped to the
\emph{true} location.

For this reason, we introduce the \emph{best mapping problem}, where we
search not the true, but the \emph{best} location of the DNA fragment,
\textit{i.e.} the sequence of the genome that is the most similar to the
read.

There are exact solutions to the best mapping problem, but the algorithms
are too slow to process the large amount of data generated by modern
sequencers. Instead, one uses heuristic methods, \textit{i.e.} algorithms
that run fast, but are not guaranteed to return the correct answer. In the
context of the best mapping problem, there are thus three possible
outcomes: $i.$ the DNA fragment is mapped to the best location, which is
called a \emph{true positive} (or an \emph{on-target hit}), $ii.$ the DNA
fragment is mapped to another location, which is called a \emph{false
positive} (or an \emph{off-target hit}), $iii.$ the DNA fragment is not
mapped at all, which is called a \emph{false negative}. A \emph{true
negative} would designate the outcome where a DNA fragment foreign to the
genome is not mapped at all, but here we assume that all DNA fragments
belong to the genome of interest.

In a seeding heuristic, one uses exact or near exact matches between the
read and the genome to collect a set of candidate locations. The read is
only compared to the sequences at the candidate locations, but not to the
rest of the genome. The advantage of this method is that there are few
candidates and that they can be discovered fast. However, the best hit may
not be in the candidate set, in which case it cannot be discovered
(\textit{i.e.}, the outcome is either a false positive or a false
negative).

A \emph{seed} designates a subsequence of fixed length that is shared
between the read and the genome. In other words, a seed is a perfect match
between a part of the read an a genomic site. The candidates consist of
the genomic sites that contain one or more seeds (this depends on the
specific heuristic).

In the theory developed below, we will assume that the \emph{true}
location of the read is also the \emph{best}. This is not always the case,
but it is unfeasible to model the behavior of the best hit in the present
context. We will later discuss how this relates to the probability that
the best hit has a seed.

\subsection{Maximal Exact Matches (MEMs)}

Here we consider a seeding variant where seeds do not have a fixed
length. Let us start by introducing the definition of a Maximal Exact
Match.

\begin{definition}
A Maximal Exact Match (MEM) is a subsequence of a read that is present in
the genome and that cannot be extended left or right --- either because
the read ends or because the extended subsequence is not in the genome.
\end{definition}

This definition presents a computational challenge: to know if a given
subsequence is a MEM, we need to know if it exists somewhere in the
genome. This is a non trivial problem in itself, but fortunately there
exists practical methods to answer the question, even for very large
genomes. Here on, we will assume that we always know which subsequences of
the read are MEMs.

A practical consideration is that short matches between the read and the
genome are irrelevant. For instance, $99.7\%$ of all the possible 12-mers
are present in the reference human genome, so an exact match of size 12 or
lower says nothing about the location of the DNA fragment. In what
follows, we will simply disregard short matches. More specifically, we
will assume that subsequences of the read can only match the original DNA
sequence or its duplicates, but not the rest of the genome. This will
simplify the discussion without loss of generality, because short
uninformative matches are ignored in all practical applications.

Fig.~\ref{fig:MEM_example} shows a concrete example of the relationship
between read errors, on-target matches, off-target matches and MEMs. The
sequence of the original DNA fragment is the first genomic hit (this is
the \emph{true} location). The second, third and fourth genomic hits are
duplicates. The mismatches between the genome and the read are
highlighted in bold. The read has three errors indicated by stars, so
\texttt{T}, \texttt{T} and \texttt{C} should actually be \texttt{A},
\texttt{A} and \texttt{G} (because the first hit is the \emph{true}
sequence).

\begin{figure}[h]
\centering
\includegraphics[scale=1]{MEM_example.pdf}
\caption{\textbf{Errors, matches and MEMs.}
Some legend.}
\label{fig:MEM_example}
\end{figure}

The MEMs are indicated by grey boxes. Technically, MEMs are subsequences
of the read, but they are shown on the genomic hits to highlight where
they match. Two MEMs can overlap, in which case they must match distinct
sequences of the genome (this follows from the definition). The last MEM
corresponds to two matches of length 4, with the first and the second
genomic hits. Such MEMs with several matches are referred to as
\emph{shared}. MEMs with a single match in the genome are referred to as
\emph{strict}.

As the name suggests, the principle of MEM seeding is to use MEM as seeds
to identify candidate genomic locations. Because short matches are
irrelevant, MEM seeds are usually requested to have a length greater than
or above a certain threshold denoted $\gamma$. We will assume that a
single MEM seed is sufficient to consider that a genomic hit is a
candidate location, which is why we will focus on the case that the read
does not have a single on-target MEM seed.

%In the example of Fig.~\ref{fig:MEM_example}, there is no MEM that matches
%the third genomic hit. It is important to understand that this can also
%happen to the best hit, as in the example shown in
%Fig~\ref{fig:MEM_eclipse}. The undesirable consequence is that some reads
%are bound to yield false positives, even if we tune the MEM seeding
%heuristic by changing the minimum seed length $\gamma$. We will refer to
%the case that the best hit is masked by off-target MEMs as \emph{MEM
%eclipse}. 
%
%\begin{figure}[h]
%\centering
%\includegraphics[scale=1]{MEM_eclipse.pdf}
%\caption{\textbf{MEM eclipse.}
%Some legend.}
%\label{fig:MEM_eclipse}
%\end{figure}
%
%If you are not already familiar with MEM seeding, we recommend that you
%take a moment to study the examples of Fig.~\ref{fig:MEM_example} and
%\ref{fig:MEM_eclipse}. This will be essential to understand the next
%sections. Before reading further, make sure that you can answer the
%following questions:
%\begin{enumerate}
%\itemsep0em
%\item What are the conditions for a MEM to be shared?
%\item What is the minimum number of duplicates for MEM eclipse?
%\end{enumerate}

\section{Combinatorial construction of reads}

In this section we give a representation of reads as combinatorial ojects.
The construction has two essential purposes: the first is to give a
generative model of reads without any on-target MEM seed, the second is to
allow us to find their weighted generating function, which in turn will
allow us to find their probability of occurrence.

\subsection{The MEM alphabet}

The nucleotide sequence of the read is not so useful to know where MEMs
are and whether they are on-target. To develop a formal method to compute
the probability that the read contains an on-target MEM seed, we will
recode reads as sequences of letters from the so-called \emph{MEM
alphabet} $\{\downarrow, \uparrow, \square\}$.

Each nucleotide of the read is recoded with one of three letters. The
$\downarrow$ symbol indicates that the nucleotide is a sequencing error.
By definition, it is a mismatch for the target. The $\uparrow$ symbol
characterizes the presence of an on-target MEM in a way that we are about
to clarify. All the other nucleotides are represented by the $\square$
symbol.

To understand the meaning of the $\uparrow$ symbol, consider the following
situation: at the left end of the read, as we try to extend the first
match as much as possible to the right, the number of hits in the genome
keeps decreasing. It may be the case that at some point, we are left with
a single hit that is on-target. The very first nucleotid where that
happens is called the \emph{pivot} and is labelled with the symbol
$\uparrow$.

Fig.~\ref{fig:sketch_MEM} shows the read of Fig.~\ref{fig:MEM_example} as
encoded with the MEM alphabet. Below the read are the matches between the
read and the genomic hit, as they are in Fig.~\ref{fig:MEM_example}. A
black square means that the nucleotide is a mismatch for the genomic
sequence, and a white or grey square means that it is a match. Grey
squares are used to highlight MEMs, as in Fig.~\ref{fig:MEM_example}.

\begin{figure}[h]
\centering
\includegraphics[scale=.85]{sketch_MEM.pdf}
\caption{\textbf{Reads and MEM seeds}. 
Some legend.}
\label{fig:sketch_MEM}
\end{figure}

Observe that the two on-target MEMs contain a pivot, but the shared MEM
does not. This is always the case. Recall that the pivot indicates that
the true location is the \emph{only} local match, so the $\uparrow$ symbol
cannot appear in on-target MEM seeds when they are shared.

Strict on-target MEM seeds are easy to find when reads are represented in
the MEM alphabet: they are the longest stretches of symbols containing the
$\uparrow$ symbol and not containing the $\downarrow$ symbol. Shared
on-target MEM seeds are trickier. They do not directly appear in this
representation so we will have to keep that in mind throughout the theory.
It is in a sense justified to put more emphasis on strict MEM seeds
because they guarantee that the true location will be found. Shared MEM
seeds can have thousands of hits in the genome in practice, so the user
may not test them all. That said, the theory developed here fully takes
shared MEM seeds into account, even if they are not visible in the
representation of the MEM alphabet.

The point of the MEM seeding alphabet is that reads can be seen as
sequences of letters, but also as sequences of \emph{segments} delimited
by the $\uparrow$ and the $\downarrow$ symbols. Because such segments are
intimately linked to the presence of on-target MEM seeds, we can give a
construction of reads that do not contain any.

\begin{definition}
\label{def:segment}
A segment is a sequence of $\square$ symbols terminated by either a
$\uparrow$ symbol or by a $\downarrow$ symbol. The unterminated sequence
of $\square$ symbols at the end of the read is called the tail --- which
can thus be an empty sequence of symbols.
\end{definition}

Following definition~\ref{def:segment}, the $\uparrow$ and $\downarrow$
symbols will be referrred to as \emph{terminators}.
Fig.~\ref{fig:sketch_segment} shows the decomposition of the read from
Fig.~\ref{fig:sketch_MEM} in segments. The size of each segment (the
number of nucleotides that compose it) is indicated below.

\begin{figure}[h]
\centering
\includegraphics[scale=.85]{sketch_segments.pdf}
\caption{\textbf{Reads as sequences of segments}. 
Some legend.}
\label{fig:sketch_segment}
\end{figure}

\subsection{The extended MEM alphabet}

An issue with the MEM alphabet is that the probabilities of the symbols
are difficult to evaluate. The $\downarrow$ symbol coincides with read
errors, which is straightforward to model. The $\uparrow$ symbol coincides
with pivots, which have a more complex behavior. We thus extend the MEM
alphabet so that the occurrence of the $\uparrow$ becomes easier to
evaluate.

Only the terminators are extended; the $\square$ symbols remain unchanged.
Let us denote the number of duplicates of the target by $N$. The
$\downarrow$ symbol is replaced by the $N+1$ symbols
$\downarrow_{/m}$, where $m$ is the number of duplicates that
\emph{match} the erroneous nucleotide $(0 \leq m \leq N)$. The $\uparrow$
symbol is replaced by the symbols $\uparrow_{/j}$, where $i
\geq 0$ is the size of the segment (including the
$\uparrow_{/j}$ symbol). Finally, we add the special terminator
$|$ that marks the end of the read. In other words, tails are considered
normal segments terminated by the symbol $|$ (with size 0).

It is important to highlight that the numbers that decorate the symbols
have a very different nature. In the case of the $\downarrow$ symbol, it
is an information about the number of sequences matching the read; in the
case of the $\uparrow$ symbol, it is the position of the pivot relative to
the previous terminator (or to the beginning of the read).

Fig.~\ref{fig:sketch_extended} shows the decomposition of the read from
Fig.~\ref{fig:sketch_segment} in segments using the extended MEM alphabet.
The numbers decorating the symbols are shown above them to avoid crowding
the figure.

\begin{figure}[h]
\centering
\includegraphics[scale=.85]{sketch_extended.pdf}
\caption{\textbf{Reads as sequences of segments in the extended MEM
alphabet}. 
Some legend.}
\label{fig:sketch_extended}
\end{figure}

There is an infinite amount of $\uparrow_{/j}$ symbols in the
extended MEM alphabet, but if the read has no on-target MEM seed, then
only the $\uparrow_{/j}$ symbols where $1 \leq i \leq \gamma-1$
can be used (if a pivot terminates a segment of size greater than or equal
to $\gamma$, then the read contains a strict on-target MEM seed). The
extended MEM alphabet is made in such a way that the numbers decorating
the $\uparrow$ symbol will allow us to exclude strict on-target MEM seeds,
and the numbers decorating the $\downarrow$ symbol will help us exclude
shared on-target MEM seeds.

\section{Symbolic and analytic representations}

\subsection{Weighted generating functions}

Let $\mathcal{A}$ be a set of combinatorial objects which have a
\emph{probability of occurrence}, and a \emph{size} that is an integer
number. The function defined as $A(z) = \sum_{k=0}^\infty a_kz^k$, where
$a_k$ is the total probability of objects of size $k$, is called the
\emph{weighted generating function} of the combinatorial objects.

In our case, we are interested in the weighted generating function of
reads without on-target MEM seed, where the coefficient $a_k$ is the
probability that a read of size $k$ has no on-target MEM seed. As $k$
becomes larger, it becomes more difficult to evaluate $a_k$, but the
combinatorial construction developed above will allow us to do this by
using established strategies involving weighted generating functions.

\subsection{The symbolic method}

The symbolic method is a strategy belonging to the general \emph{analytic
combinatorics} framework. The idea is to combine simple combinatorial
objects into more complex objects. Each combinatorial operation on the
objects corresponds to a mathematical operation on their weighted
generating functions. One can thus obtain the weighted generating function
of complex objects, and from there extract their probability of
occurrence.

The combinatorial construction of reads using the extended MEM seeding
alphabet gives us a method to express their weighted generating function
from that of segments. More specifically, we can describe complex
sequences of segments through a so-called \emph{transfer matrix} $M(z)$
that specifies which segments can follow one another in the reads of
interest.

The segments types are identified by their terminators and arranged in a
predefined oder. The entry of $M(z)$ at coordinates $(i,j)$ is the
weighted generating function of segments with the $j$-th terminator that
can be inserted immediately after segments with the $i$-th terminator. The
matrix $M(z) + M(z)^2 + \ldots = M(z) \cdot (I-M(z))^{-1}$ contains the
weighted generating functions of the reads of all sizes. The entry that
corresponds to terminators $\downarrow_{/0}$ and $|$ is the weighted
generating function of the reads of interest (the beginning of the read is
equivalent to a ghost segment terminated by $\downarrow_{/0}$,
\textit{i.e.} a mismatch for all the sequences).

With the transfer matrix $M(z)$, with thus have a way of computing the
weighted generating function of the reads without on-target MEM seed, and
from there we can extract teir probability of occurrence. Our next task is
thus to specify $M(z)$.

@@

\subsection{Match streak and masks}

For each sequence, we define the \emph{match streak} at a given nucleotide
as the number of nucleotides since the last mismatch on the left. For
instance, in Fig.~\ref{fig:MEM_example}, the match streak of the first
genomic hit at the ninth nucleotide is $1$ while for the second genomic
hit, it is $7$.

\begin{definition}
At a given position of the read, a duplicate sequence is a \emph{hard
mask} if its success streak is strictly longer than the match streak of
the target. A duplicate thread is a \emph{soft mask} if its has the same
match streak as the target.
\end{definition}

For instance, the second genomic hit in the example above is a hard mask
at the ninth nucleotide. The point of hard and soft masks is that they are
directly linked with MEMs. A nucleotide with a hard mask cannot belong to
an on-target MEM. A nucleotide without mask belongs to a strict on-target
MEM. A nucleotide with a soft mask may belong to a strict on-target MEM,
to a shared on-target MEM, or may not belong to any MEM. The distinction
between the three cases will be clarified in the specification of the
transfer matrix.

The extended MEM alphabet is nothing more than a way to keep track of the
hard and soft masks. For instance, a pivot (associated with the symbol
$\uparrow$) is a nucleotide where the number of masks becomes is reduced
to 0. If there are $N$ duplicates, the symbol $\downarrow_{/m}$ indicates
that there $m$ hard masks and $N-m$ soft masks.

\subsection{Error model and divergence of the duplicates}

We assume that the target sequence has $N$ duplicates. We further assume
that duplication was instantaenous and that all $N+1$ sequences diverge
independently from each other at a constant rate. In other words, we
ignore the complications due to the genealogy of the duplication events
and we simply assume that at each position, any given duplicate is
identical to the target with probability $1-\mu$. If it is not, we
assume that the duplicate has any of the remaining three nucleotides with
equal probability (\textit{i.e.} each is found with probability $\mu/3$).

We also assume that the sequencing instrument has a constant substitution
rate $p$, and that insertions and deletions never occur. In case of a
substitution, we again assume that the remaining three nucleotides are
equally likely.

On a read error the target is never identical to the read (because we
assume that the target is the true sequence), and a duplicate is identical
to the read with probability $mu/3$ (the probability that the duplicate is
different from the target, times the probability that the nucleotide is
the same as the error). On a correct nucleotide, the target is always
identical to the read, and a dupicate is identical with probability
$1-\mu$ (the probablity that the duplicate is identical to the target).

With these assumptions, we are now ready to describe the entries of the
transfer matrix $M(z)$, which we do in order of increasing difficulty.

\subsection{Segments following $\uparrow_{/j}$}

A $\uparrow_{/j}$ symbol marks a pivot and the absence of masks. The
segment is thus the beginning of an on-target MEM of size at least $j$.
The MEM will extend until the next sequencing error, or until the end of
the read, so the number of $\square$ symbols in the next segment must be
at most $\gamma-j-1$ and it must be terminated by a $\downarrow$ symbol or
by the tail terminator $|$.

In this case, the matches between the read and the duplicates are
irrelevant until the terminator. The weighted genearting function of each
$\square$ symbol is simply $qz$ ($q$ is the probability that the
nucleotide is not an error and $z$ marks objects of size $1$).

\begin{definition}
The probability that a symbol is $\downarrow_{/m}$ given that the
nucleotide is a read error is
\begin{equation}
\label{eq:omega}
\omega_m = {N \choose m} \big(1 - \mu/3\big)^{N-m} \big(\mu/3\big)^m.
\end{equation}
\end{definition}

The weighted generating function of the $\downarrow_{/m}$ terminator is
thus $\omega_m pz$ (the probability of the $\downarrow_{/m}$ is $\omega_m
p$ and $z$ marks objects of size $1$).

The entries of the transfer matrix that correspond to transitions from
$\uparrow_{/j}$ to other segments are now easy to find. The weighted
generating function of the segments terminated by $\downarrow_{/m}$
following a segment terminated by $\uparrow_{/j}$ is
\begin{equation}
\label{eq:D}
D_{j,m}(z) = \omega_m pz \sum_{i=0}^{\gamma-j-1} (qz)^i.
\end{equation}

And the weighted generating function of the tail segments following
segments terminated by $\uparrow_{/j}$ is
\begin{equation}
\label{eq:E}
E_j(z) = \sum_{i=0}^{\gamma-j-1} (qz)^i.
\end{equation}

\subsection{Segments following $\downarrow_{/m}$}

At a $\downarrow_{/m}$ symbol, there are $m$ hard masks and $N-m$ soft
masks. If all the masks disappear before the first read error, the next
terminator will be a $\uparrow$ symbol, otherwise it will be a
$\downarrow$ symbol.

\subsubsection*{Case 1: the terminator is $\uparrow_{/i}$}

\begin{definition}
The probability that a given duplicate sequence contains a mismatch in a
sequence of $i$ error-free nucleotides is
\begin{equation}
\label{eq:xi}
\xi_i = 1-(1-\mu)^i.
\end{equation}
\end{definition}

With this notation, the probability that at least one mask survives a
sequence of $i$ error-free nucleotides is thus $1-\xi_i^N$, and the
probability that there remains a mask at the $i-1$-th but not at the 
$i$-th error-free nucleotide is $\xi_i^N - \xi_{i-1}^N$. From this we
conclude that the weighted generating function of the segments terminated
by $\uparrow_{/i}$ following a segment terminated by $\downarrow_{/m}$ is
\begin{equation}
\label{eq:B}
B_i(z) = \Big( \xi_{i-1}^N-\xi_i^N \Big) (qz)^i.
\end{equation}

In the equation above, $q^i$ is the probability that there is no
sequencing error in $i$ nucleotides and $z^i$ marks objects of size $i$.
As mentioned above, $i < \gamma$ for reads without on-target MEM seed.
Also note that this expression is the same for all $\downarrow$ symbols;
it does not depend on $m$.

\subsubsection*{Case 2a: the terminator $|$ comes before the $\gamma$-th
nucleotide}

In this case there can be no on-target MEM seed. We must just enforce the
condition that at least one of the $N$ masks survives until the end of the
segment in order to exclude the $\uparrow$ symbol. The weighted generating
function is
\begin{equation*}
\sum_{i=0}^{\gamma-1} \Big(1 - \xi_i^N \Big) (qz)^i.
\end{equation*}

\subsubsection*{Case 2b: the terminator $|$ comes after the $\gamma$-th
nucleotide}

In this case, the soft masks are irrelevant. Even if they survive until
the end of the segment, there will be a shared on-target MEM seed. So we
must enforce the condition that at least one hard mask survives until the
end of the segment (which is impossible if $m = 0$).
\begin{equation*}
\sum_{i=\gamma}^\infty \Big(1 - \xi_i^m \Big) (qz)^i.
\end{equation*}

The transition from $\downarrow_{/m}$ to $|$ is
\begin{equation}
\label{eq:C}
C_m(z) =
\sum_{i=0}^{\gamma-1} \Big(1 - \xi_i^N \Big) (qz)^i +
  \sum_{i=\gamma}^\infty \Big(1 - \xi_i^m \Big) (qz)^i.
\end{equation}

\subsubsection*{Case 3a: the terminator $\downarrow_{/n}$ comes before the
$\gamma$-th nucleotide}
